---
title: "Visual Introduction to Maximum Likelihood Estimation"
author: "Jason Bryer, Ph.D."
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
    code_folding: hide
vignette: >
  %\VignetteIndexEntry{Visual Introduction to Maximum Likelihood Estimation}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, error = FALSE,
					  fig.align = 'center')

set.seed(2112)

library(tidyverse)
library(ggplot2)
library(reshape2)
library(gganimate)
library(magick)
library(cowplot)
library(DT)

data(mtcars)
```

[Maximum Likelihood Estimation](https://en.wikipedia.org/wiki/Maximum_likelihood_estimation) (MLE) is an important procedure for estimating parameters in statistical models. It is often first encountered when modeling a dichotomous outcome variable vis-à-vis logistic regression. However, it is the backbone of [generalized linear models](https://en.wikipedia.org/wiki/Generalized_linear_model) (GLM) which allow for error distribution models other than the normal distribution. This document was inspired by the [blog post by Jorge Cimentada](https://cimentadaj.github.io/blog/2020-11-26-maximum-likelihood-distilled/maximum-likelihood-distilled/) to provide some additional details about each step of MLE.

We will begin with a typical bivariate regression using the `mtcars` data set where we wish to predict `mpg` (miles per gallon) from `wt` (weight in 1,000 lbs). Figure 1 is a scatter plot showing the relationship between these two variables.


```{r, echo = FALSE, fig.cap = 'Figure 1. Scatter plot of weight versus miles per gallan.'}
ggplot(mtcars, aes(x = wt, y = mpg)) + geom_point()
```

Our goal is to estimate

$$Y_{mpg} = \beta_{wt} X + e$$
where $\beta_{wt}$ is the slope and $e$ is the intercept.


## Ordinary Least Squares

With ordinary least squares (OLS) regression our goal is to minimize the residual sum of squares (RSS):

$$RSS=\sum^{n}_{i=1} \left( y_{i}-f(x_{i})\right)^{2}$$

where $y_i$ is the variable to be predicted, $f(x_i)$ is the predicted value of $y_i$, and $n$ is the sample size. Here is what we know about regression:

* The correlation measures the strength of the relationship between x and y (see [this shiny app](https://shiny.rit.albany.edu/stat/rectangles/) for an excellent visual overview of correlations).
* The correlation ranges between -1 and 1.
* The mean of x and y must fall on the line.
* The slope is the change in y over the change in x and is defined as $m = r * (\sigma_y / \sigma_x)$. Where $r$ is the correlation and $\sigma$ is the sample standard deviation.

We can easily calculate the RSS for various correlations ($r$) ranging between -1 and 1. Figure 2 visualizes the RSS.

```{r class.source = 'fold-show'}
y <- mtcars$mpg
x <- mtcars$wt
mean.y <- mean(y)
mean.x <- mean(x)
sd.y <- sd(y)
sd.x <- sd(x)
ols <- tibble(
	r = seq(-1, 1, by = 0.025),            # Correlation
	m = r * (sd.y / sd.x),                 # Slope
	b = mean.y - m * mean.x                # Intercept
) %>% rowwise() %>%
	mutate(ss = sum((y - (m * x + b))^2)) %>% # Sum of squares residuals
	as.data.frame()
datatable(ols)
```

```{r, echo = FALSE, fig.cap = 'Figure 2. Residual sum of squares.'}
ggplot(ols, aes(x = r, y = ss)) + geom_path() + geom_point()
```

The correlation with the correlation the resulted in the smallest RSS is `r ols %>% filter(ss == min(ss)) %>% select(r) %>% unname()`.

```{r class.source = 'fold-show'}
ols %>% filter(ss == min(ss)) # Select the row with the smallest sum of squares residuals
```

Calculating the correlation in R gives us `r cor(x, y)` and the slope is `r cor(x, y) * (sd.y / sd.x)` which is close to our estimate here. We could get a more accurate result if we tried smaller steps in the correlation (see the `by` parameter in the `seq` function above).

## Minimizing RSS Algorithmically

This approach works well here because the correlation is bounded between -1 and 1 and we can easily calculate the RSS for a bunch of possible correlations. However, there are more efficient ways of finding the correlation that minimizes the RSS than trying correlations equally distributed across the range. For example, consider the following algorithm:

1. Calculate the RSS for $r = 0$. 
2. Calculate the RSS for $r = 0.5$ If $RSS_{0.5} < RSS_{0}$ then calculate the RSS with $r = 0.75$, else calculate the RSS with $r = -0.5%$

We can repeat this procedure, essentially halving the distance in each iteration until we find a sufficiently small RSS. This process is, in essence, the idea of numerical optimization procedures. In R, the `optim` function implements the [Nedler-Mead](https://en.wikipedia.org/wiki/Nelder–Mead_method) method for optimizing a set of parameters. The details of *how* the algorithm works is beyond the scope of this article (see this [interactive tutoral](https://www.benfrederickson.com/numerical-optimization/) by Ben Frederickson for a good introduction), instead we will focus on *what* the algorithm does. To begin, we must define a function that calculates a metric for which the optimizer is going to minimize (or maximize). Let's start with RSS:

```{r class.source = 'fold-show'}
# Change to residSumSquares
sumsquares <- function(parameters, predictor, outcome) {
	a <- parameters[1] # Intercept
	b <- parameters[2] # beta coefficient
	predicted <- a + b * predictor
	residuals <- outcome - predicted
	ss <- sum(residuals^2)
	return(ss)
}
```

The `parameters` is a vector of the parameters the algorithm is going to minimize (or maximize). Here, these will be the slope and intercept. The `predictor` and `outcome` are parameters passed through from the `...` parameter on the `optim` function and are necessary for us to calculate the RSS. We can now get the RSS for any set of parameters.

```{r class.source = 'fold-show'}
sumsquares(c(37, -5), mtcars$wt, mtcars$mpg)
```

**Small Digression** In order to explore each step of the algorithm, we need to wrap the `optim` function to capture the parameters and output of the function. This function will overwrite the `optim` function in the current environment to add to elements to the returned list: `iterations` is the raw list of the parameters and output saved and `iterations_df` is a `data.frame` version. Typically overwriting a function like this is bad practice (please don't email me) but is efficient for my purposes here.

```{r class.source = 'fold-show'}
optim_save <- function(par, fn, ...) {
	iterations <- list()
	wrap_fun <- function(parameters, ...) {
		n <- length(iterations)
		result <- fn(parameters, ...)
		iterations[[n + 1]] <<- c(parameters, result)
		return(result)
	}
	optim_out <- stats::optim(par, wrap_fun, ...)
	optim_out$iterations <- iterations
	optim_out$iterations_df <- as.data.frame(do.call(rbind, iterations))
	names(optim_out$iterations_df) <- c(paste0('Param', 1:length(par)), 'Result')
	optim_out$iterations_df$Iteration <- 1:nrow(optim_out$iterations_df)
	return(optim_out)
}
```


We can now call the `optim_save` function with our `sumsquares` function. We initialize the algorithm with two random values for the intercept and slope, respectively. Note that we are using Broyden, Fletcher, Goldfarb, and Shanno optimization method which allows for the specification of bounds on the parameter estimates which we will use later.

```{r class.source = 'fold-show'}
optim.rss <- optim_save(
	par = runif(2),
	fn = sumsquares, 
	method = "L-BFGS-B",
	predictor = mtcars$wt,
	outcome = mtcars$mpg
)
```

The `par` parameter provides the final parameter estimates.

```{r class.source = 'fold-show'}
optim.rss$par
```

We can see that the parameters are accurate to at least four decimal places to the OLS method used by the `lm` function.

```{r class.source = 'fold-show'}
lm.out <- lm(mpg ~ wt, data = mtcars)
lm.out$coefficients
```

It took the `optim` function `r nrow(optim.rss$iterations_df)` iterations to find the optimal set of parameters that minimized the RSS. Figure 3 shows the value of the parameters (i.e. intercept and slope) and the RSS for each iteration.

```{r, fig.cap = 'Figure 3. Output of the optimizaiton procedure at each iteration.'}
df <- optim.rss$iterations_df
names(df) <- c('Intercept', 'Slope', 'SumSquares', 'Iteration')
df %>% melt(id.var = 'Iteration') %>%
	ggplot(aes(x = Iteration, y = value, color = variable)) +
	geom_point(size = 1) + geom_path() +
	facet_wrap(~ variable, scales = "free_y", ncol = 1) +
	xlab('Iteration') + ylab('') + theme(legend.position = 'none')
```

## Likelihood

Now that we have laid the groundwork for finding parameters algorithmically, we need to introduce another way of evaluating how well parameters *fit* the data, namely the likelihood. First, let's revisit what we are doing in OLS. Figure 4 is a scatter plot of our observations, the OLS regression line in blue, and one observation highlighted in red with the residual as a red line. With OLS, we square the residual for every observation, thereby making all values positive, and summing them. There is, however, another way of estimating fit that doesn't rely on the residuals.

```{r, fig.cap='Figure 4. Scatter plot with residuals for one observation.'}
pt <- 1 # Which observation do we want to explore
mtcars$fitted_mpg <- fitted(lm.out)
a <- lm.out$coefficients[1]
b <- lm.out$coefficients[2]
sigma <- summary(lm.out)$sigma
fitted.pt <- mtcars[pt,] * a + b
ggplot(mtcars, aes(x = wt, y = mpg)) +
	geom_point() +
	geom_segment(data = mtcars[pt,], color = 'red', size = 1,
				 aes(x = wt, xend = wt, y = mpg, yend = fitted_mpg)) +
	geom_point(data = mtcars[pt,], color = 'red', size = 4) +
	geom_smooth(method = 'lm', formula = y ~ x, se = FALSE)
```


We often think of probabilities as the areas under a fixed distribution. For example, the first car in `mtcars` is `r row.names(mtcars)[pt]` with an average miles per gallon of `r mtcars[pt,]$mpg` and weighs `r mtcars[pt,]$wt * 1000`lbs. The probability of a car with a miles per gallon less than `r row.names(mtcars)[pt]` given the data we have in `mtcars` is `r pnorm(mtcars[pt,]$mpg, mean(mtcars$mpg), sd = sd(mtcars$mpg))` and is depicted in Figure 5.

```{r, fig.cap = 'Figure 5. Probability distribution of miles per gallan.'}
ggplot() +
	stat_function(fun = dnorm, n = 101, geom = "line",
				  args = list(mean = mean(mtcars$mpg),
				  			  sd = sd(mtcars$mpg))) +
	stat_function(fun = dnorm, n = 101, geom = "area", fill = "steelblue",
				  args = list(mean = mean(mtcars$mpg),
				  			sd = sd(mtcars$mpg)),
				  xlim = c(mean(mtcars$mpg) - 3 * sd(mtcars$mpg), mtcars[pt,]$mpg)) +
	geom_segment(aes(x = mtcars[pt,]$mpg, xend = mtcars[pt,]$mpg),
				 y = 0, yend = dnorm(y[pt], mean(mtcars$mpg), sd(mtcars$mpg))) +
	xlim(mean(mtcars$mpg) - 3 * sd(mtcars$mpg), mean(mtcars$mpg) + 3 * sd(mtcars$mpg)) +
	xlab('Miles Per Gallon') + ylab('Density')
```

For probabilities, we are working with a fixed distribution, that is:

$$pr(data\ |\ distribution)$$
The likelihood are the y-axis values (i.e. density) for fixed data points with distributions that can move, that is: 

$$L(distribution\ |\ data)$$
The likelihood is the height of the density function. Figure 6 depicts two likelihood for two observations. The mean of each distribution is equal to $\beta_{wt} X + e$ and the intercept (also known as the error term) defines the standard deviation of the distribution.

```{r, fig.cap = 'Figure 6. Likelihood of a car having the observed mpg given the model parameters for two observations.'}
pt1 <- 1
p1 <- ggplot() +
	stat_function(fun = dnorm, n = 101,
				  args = list(mean = a + b * mtcars[pt1,]$wt,
				  			  sd = sigma)) +
	geom_segment(aes(x = mtcars[pt1,]$mpg, xend = mtcars[pt1,]$mpg),
				     y = 0, yend = dnorm(y[pt1], a + b * x[pt1], sigma)) +
	geom_point(aes(x = mtcars[pt1,]$mpg, y = dnorm(y[pt1], a + b * x[pt1], sigma)),
			   color = 'red', size = 4) +
	xlim(mean(y) - 3 * sd(y), mean(y) + 3 * sd(y)) +
	xlab('') + ylab('Density')
pt2 <- 5
p2 <- ggplot() +
	stat_function(fun = dnorm, n = 101,
				  args = list(mean = a + b * mtcars[pt2,]$wt,
				  			  sd = sigma)) +
	geom_segment(aes(x = mtcars[pt2,]$mpg, xend = mtcars[pt2,]$mpg),
				     y = 0, yend = dnorm(y[pt2], a + b * x[pt2], sigma)) +
	geom_point(aes(x = mtcars[pt2,]$mpg, y = dnorm(y[pt2], a + b * x[pt2], sigma)),
			   color = 'red', size = 4) +
	xlim(mean(y) - 3 * sd(y), mean(y) + 3 * sd(y)) +
	# xlim((a + b * x[pt2]) - 3 * sigma, (a + b * x[pt2]) + 3 * sigma) +
	xlab('Miles per Gallon') + ylab('Density')
plot_grid(p1, p2, ncol = 1)
```

We can then calculate the likelihood for each observation in our data. Unlike OLS, we now want to *maximize* the sum of these values. Also, we are going to use the log of the likelihood so we can add them instead of multiplying. We can now define our log likelihood function:

```{r class.source = 'fold-show'}
# TODO: Rename error to maybe root mean squared error (sd of the residuals?)
loglikelihood <- function(parameters, predictor, outcome) {
	a <- parameters[1]     # intercept
	b <- parameters[2]     # slope / beta coefficient
	sigma <- parameters[3] # error
	ll.vec <- dnorm(outcome, a + b * predictor, sigma, log = TRUE)
	return(sum(ll.vec))
}
```

Note that we have to estimate a third parameter, sigma, which is the error term and defines the standard deviation for the normal distribution for estimating the likelihood. We can now calculate the log-likelihood for any combination of parameters.

```{r class.source = 'fold-show'}
loglikelihood(c(37, -5, sd(mtcars$mpg)),
			  predictor = mtcars$wt,
			  outcome = mtcars$mpg)
```

## Maximum Likelihood Estimation

We can now use the `optim_save` function to find the parameters that *maximize* the log-likelhood. Note two important parameter changes:

1. We are specifying the `lower` parameter so that the algorithm will not try negative values for sigma since the variance cannont be negative.
2. The value for the `control` parameter indicates that we wish to maximize the values instead of minimizing (which is the default).

```{r class.source = 'fold-show'}
optim.ll <- optim_save(
	runif(3),                     # Random initial values
	loglikelihood,                # Log-likelihood function
	lower = c(-Inf, -Inf, 1.e-5), # The lower bounds for the values, note sigma (error), cannot be negative
	method = "L-BFGS-B",
	control = list(fnscale = -1), # Indicates that the maximum is desired rather than the minimum
	predictor = mtcars$wt,
	outcome = mtcars$mpg
)
```

We can get our results and compare them to the results of the `lm` function and find that they match to at least four decimal places.

```{r class.source = 'fold-show'}
optim.ll$par[1:2]
lm.out$coefficients
```

Figure 7 shows the estimated regression line for each iteration of the optimization procedure (on the left; OLS regression line in blue; MLE regression line in black) with the estimated parameters and log-likelihood for all iterations on the left.

```{r, cache = TRUE, fig.cap = 'Figure 7. Animation of parameter estimates for each iteration of the optimization procedure.'}
df <- optim.ll$iterations_df
names(df) <- c('Intercept', 'Slope', 'Sigma', 'LogLikelihood', 'Iteration')
p1 <- ggplot(mtcars, aes(x = wt, y = mpg)) +
	geom_smooth(method = lm, formula = y ~ x, se = FALSE) +
	geom_abline(data = df, aes(intercept = Intercept, slope = Slope)) +
	geom_point(data = mtcars, aes(x = wt, y = mpg)) +
	transition_time(Iteration) +
	labs(title = "Iteration: {frame_time}") +
	shadow_wake(wake_length = 0.1, alpha = FALSE)
p1_gif <- animate(p1, width = 480, height = 480)

df.melt <- df %>% melt(id.var = 'Iteration')
p2 <- ggplot(df.melt, aes(x = Iteration, y = value, color = variable)) +
	geom_vline(data = data.frame(Iteration2 = df$Iteration),
			   aes(xintercept = Iteration2, frame = Iteration2)) +
	geom_path() +
	facet_wrap(~ variable, scales = "free_y", ncol = 1) +
	xlab('Iteration') + ylab('Parameter Estimate') +
	transition_time(Iteration2)
p2_gif <- animate(p2, width = 480, height = 480)

new_gif <- image_append(c(p1_gif[1], p2_gif[1]))
for(i in 2:100){
	combined <- image_append(c(p1_gif[i], p2_gif[i]))
	new_gif <- c(new_gif, combined)
}
new_gif
```

Figure 8 superimposes the normal distribution from which the log-likelihood is determined. The distribution is centered on $\hat{y}$. The height of the distribution (i.e. density) at $y$ is the likelihood. We take the log of this value to get the log-likelihood. These log-likelihoods are calculated for each observation and summed. Maximum likelhood estimation is attempting to find the parameters (i.e. slope and intercept) that maximizes the log-likelhood.

```{r, fig.cap = 'Figure 8. Likelihood for one observeration superimposed on scatter plot.'}
intercept <- optim.ll$par[1]
slope <- optim.ll$par[2]
sigma <- optim.ll$par[3]
pt <- 2

heightFactor <- 1 # Increase the multiplier to make the distribution higher
k <- 4 # How many standard deviations to plot
x <- seq(-k * sigma, k * sigma, length.out = 50)
y <- dnorm(x, 0, sigma) / dnorm(0, 0, sigma) * heightFactor
x0 <- mtcars[pt,]$wt
y0 <- slope * x0 + intercept
path <- data.frame(x = y + x0, y = x + y0)
segment <- data.frame(x = x0, y = y0 - k*sigma, xend = x0, yend = y0 + k*sigma)
segment2 <- data.frame(x = x0, 
					   y = mtcars[pt,]$mpg, 
					   xend = dnorm(abs(y0 - mtcars[pt,]$mpg), 0, sigma) / dnorm(0, 0, sigma) * heightFactor + x0, 
					   yend = mtcars[pt,]$mpg)

ggplot(mtcars, aes(x = wt, y = mpg)) +
	# geom_smooth(method = lm, formula = y ~ x, se = FALSE) +
	geom_abline(intercept = intercept, slope = slope) +
	geom_point(data = mtcars, aes(x = wt, y = mpg)) +
	geom_segment(data = segment, aes(x = x, y = y, xend = xend, yend = yend)) +
	geom_segment(data = segment2, aes(x = x, y = y, xend = xend, yend = yend), color = 'red') +
	geom_point(data = mtcars[pt,], aes(x = wt, y = mpg), color = 'red', size = 2) +
	# geom_point(data = segment2, aes(x = xend, y = y), color = 'red', size = 2) +
	geom_vline(xintercept = x0) +
	# geom_hline(yintercept = y0) + # Verify the distribution is centered on y-hat
	geom_path(data = path, aes(x,y), color = "blue")
```


Figure 9 depicts the likelihoods for the first 16 observations.

```{r, cache = TRUE, fig.cap = 'Figure 9. Likelihoods of the first 16 observations for the final parameter estimates.'}
tmp <- df %>% filter(Iteration == nrow(df))
plots <- list()
nplots <- 16 #nrow(mtcars)
for(i in 1:min(nplots, nrow(mtcars))) {
	a <- tmp[1,]$Intercept
	b <- tmp[1,]$Slope
	sigma <- tmp[1,]$Sigma
	predictor <- mtcars$wt[i]
	predicted.out <- a + b * predictor
	outcome <- mtcars$mpg[i]
	d <- dnorm(outcome, predicted.out, sigma)
	plots[[i]] <- ggplot() +
		stat_function(fun = dnorm,
					  n = 101,
					  args = list(mean = predicted.out, sd = sigma)) +
		annotate(geom = 'segment', x = outcome, y = 0, xend = outcome, yend = d, color = 'red') +
		annotate(geom = 'point', x = outcome, y = d, color = 'red', size = 2) +
		xlim(c(min(mtcars$mpg, predicted.out - 3 * sigma),
			   max(mtcars$mpg, predicted.out + 3 * sigma))) +
		ylim(c(0, .2)) +
		ylab('') + xlab(row.names(mtcars)[i])
}
plot_grid(plotlist = plots)
```



## Generalized Linear Models

Generalized linear models (GLM) are a generalization of OLS that allows for the response variables (i.e. dependent variables) to have an error distribution that is not normally distributed. All generalized linear models have the following three characteristics:

1. A probability distribution describing the outcome variable 
2. A linear model  
	$\eta = \beta_0+\beta_1 X_1 + \cdots + \beta_n X_n$
3. A link function that relates the linear model to the parameter of the outcome distribution  
	$g(p) = \eta$ or $p = g^{-1}(\eta)$

We can estimate GLMs using MLE as described above. What will change is the log-likelihood function.


## Logistic Regression

Logistic regression is a GLM used to model a binary categorical variable using numerical and categorical predictors. We assume a binomial distribution produced the outcome variable and we therefore want to model *p* the probability of success for a given set of predictors. Instead of fitting a line (or a plane for two predictors, etc. for higher dimensions) we wish to fit the data to the logistic function which is defined as:

$$ \sigma \left( t \right) =\frac { { e }^{ t } }{ { e }^{ t }+1 } =\frac { 1 }{ 1+{ e }^{ -t } }  $$

```{r, fig.cap = 'Figure 10. Logistic curve'}
logistic <- function(t) { 
	return(1 / (1 + exp(-t))) 
}
ggplot() +
	stat_function(fun = logistic, n = 101) +
	xlim(-4, 4) + xlab('x')
```


To finish specifying the Logistic model we just need to establish a reasonable link function that connects $\eta$ to $p$. There are a variety of options but the most commonly used is the logit function which is specified as:

$$logit(p) = \log\left(\frac{p}{1-p}\right),\text{ for $0\le p \le 1$}$$

We can specify *t* as a linear combination of our predictors (independent variables).

$$ t = \beta_0 + \beta_1 x $$

The logistic function can now be rewritten as:

$$ F\left( x \right) =\frac { 1 }{ 1+{ e }^{ -\left( { \beta  }_{ 0 }+\beta _{ 1 }x \right)  } } $$

Consider the following data set where we wish to predict whether a student will pass an exam based upon the number of hours they studied.

```{r class.source = 'fold-show'}
study <- data.frame(
    Hours=c(0.50,0.75,1.00,1.25,1.50,1.75,1.75,2.00,2.25,2.50,2.75,3.00,
            3.25,3.50,4.00,4.25,4.50,4.75,5.00,5.50),
    Pass=c(0,0,0,0,0,0,1,0,1,0,1,0,1,0,1,1,1,1,1,1)
)
```

```{r, fig.cap = 'Figure 11. Boxplot of hours studied by passing.'}
ggplot(study, aes(x = factor(Pass), y = Hours)) + geom_boxplot() + xlab('Pass') + ylab('Hours Studied')
```

First, we need to define logit function and the log-likelihood function that will be used by the `optim` function. Instead of using the normal distribution as above (using the `dnorm` function), we are using a binomial distribution and the logit to link the linear combination of predictors.

```{r class.source = 'fold-show'}
logit <- function(x, beta0, beta1) {
	return( 1 / (1 + exp(-beta0 - beta1 * x)) )
}
loglikelihood.binomial <- function(parameters, predictor, outcome) {
	a <- parameters[1] # Intercept
	b <- parameters[2] # beta coefficient
	p <- logit(predictor, a, b)
	ll <- sum( outcome * log(p) + (1 - outcome) * log(1 - p))
	return(ll)
}
```

Now we can call the `optim` function and get the final parameter estimates.

```{r class.source = 'fold-show'}
optim.binomial <- optim_save(
	c(0, 1), # Initial values
	loglikelihood.binomial,
	method = "L-BFGS-B",
	control = list(fnscale = -1),
	predictor = study$Hours,
	outcome = study$Pass
)
```

In R, the `glm` (short for generalized linear models) function implements logistic regression when the `family = binomial(link = 'logit')` parameter is set. See `?glm` for other families of models to estimate models with other underlying distributions. We can see that our estimate matches the results of `glm` to a rounding error.

```{r}
optim.binomial$par
lr.out <- glm(Pass ~ Hours, data = study, family = binomial(link = 'logit'))
lr.out$coefficients
```

```{r}
# Redefine the logistic function to include parameter estimates
logistic <- function(x, beta0, beta1) {
    return(1 / (1 + exp(-1 * (beta0 + beta1 * x)) ))
}

beta0 <- optim.binomial$par[1]
beta1 <- optim.binomial$par[2]

ggplot(study, aes(x = Hours, y = Pass)) +
	# geom_smooth(method = 'glm', formula = y ~ x, se = FALSE,
	# 			method.args = list(family = binomial(link = 'logit'))) +
	geom_point(aes(color = logistic(Hours, beta0, beta1) > 0.5)) +
	stat_function(fun = logistic, n = 101, 
				  args = list(beta0 = beta0, beta1 = beta1) ) +
	scale_color_hue('Predicted Pass > 0.5') +
	theme(legend.position = c(0.85, 0.15))
```

Let's explore the process of the numeric optimizer. For this model, it took `r nrow(optim.binomial$iterations_df)` iterations to converge to resulting parameters.


```{r, cache = TRUE}
df <- optim.binomial$iterations_df
names(df) <- c('Intercept', 'Hours', 'LogLikelihood', 'Iteration')
xlim <- c(0, 6) # Hard coding for now
df2 <- data.frame(Iteration = rep(1:nrow(df), each = 100))
xvals <- seq(xlim[1], xlim[2], length.out = 100)
tmp <- apply(
	df, 1, FUN = function(x) {
		logistic(xvals, x[1], x[2])
	}
) %>% as.data.frame()
names(tmp) <- 1:ncol(tmp)
tmp <- melt(tmp)
names(tmp) <- c('Iteration', 'Pass')
tmp$Hours <- rep(xvals, nrow(df))

nFrames <- nrow(df) * 2
p1 <- ggplot() + 
	geom_smooth(data = study, aes(x = Hours, y = Pass),
		method = 'glm', formula = y ~ x, se = FALSE, alpha = 0.5,
		method.args = list(family = binomial(link = 'logit'))) +
	geom_point(data = study, aes(x = Hours, y = Pass)) + 
	geom_path(data = tmp, aes(x = Hours, y = Pass, group = Iteration)) +
	transition_states(Iteration) +
	labs(title = "Iteration: {round(frame/2)}") +
	shadow_wake(wake_length = 0.1, alpha = FALSE) +
	ease_aes("cubic-in")
p1_gif <- animate(p1, nframes = nFrames, width = 480, height = 480)

df.melt <- df %>% melt(id.var = 'Iteration')
p2 <- ggplot(df.melt, aes(x = Iteration, y = value, color = variable)) +
	geom_vline(data = data.frame(Iteration2 = df$Iteration),
			   aes(xintercept = Iteration2, frame = Iteration2)) +
	geom_path() +
	facet_wrap(~ variable, scales = "free_y", ncol = 1) +
	xlab('Iteration') + ylab('Parameter Estimate') +
	theme(legend.position = 'none') +
	transition_time(Iteration2)
p2_gif <- animate(p2, nframes = nFrames, width = 480, height = 480)

new_gif <- image_append(c(p1_gif[1], p2_gif[1]))
for(i in 2:nFrames){
	combined <- image_append(c(p1_gif[i], p2_gif[i]))
	new_gif <- c(new_gif, combined)
}
new_gif
```


```{r}
pt <- 1
beta0 <- optim.binomial$par[1]
beta1 <- optim.binomial$par[2]

df2 <- data.frame(x = xvals,
				  p = logistic(xvals, beta0, beta1))

ggplot(df2, aes(x = x, y = p)) + 
	geom_path() + 
	geom_segment(aes(x = study[pt,]$Hours, xend = study[pt,]$Hours,
					 y = study[pt,]$Pass, yend = logit(study[pt,]$Hours, beta0, beta1)),
				 color = 'red', size = 1) +
	geom_point(aes(x = study[pt,]$Hours, y = logit(study[pt,]$Hours, beta0, beta1)),
			   color = 'red', size = 4)

study$p <- logit(study$Hours, beta0, beta1)
ggplot(df2, aes(x = x, y = p)) + 
	geom_path() + 
	geom_segment(data = study, aes(x = Hours, xend = Hours,
								   y = Pass, yend = p), color = 'red', size = 1) +
	geom_point(data = study, aes(x = Hours, y = p), color = 'red', size = 4)

```
